---
title: "Open Policing - Raleigh"
author: "Otho"
date:  "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

Reproduce Nadieh Bremer and Zan Armstrong famouse [Baby Spike visualization](https://www.visualcinnamon.com/portfolio/baby-spike)

```{r}
library(tidyverse)
library(rlang)
library(lubridate)
library(scales)
```


```{r}
dat_path <- "data/2-12-open-policing.Rdata"

# data for Raleigh, because I have been there
dat_url <- paste0("https://stacks.stanford.edu/",
                  "file/druid:tr137st9964/tr137st9964",
                  "_nc_raleigh_2019_02_25.csv.zip")

if(!file.exists(dat_path)) {
  # one temporary file zipped
  # and one with the csv
  temp <- tempfile()
  temp2 <- tempfile()
  download.file(dat_url, destfile = temp)
  
  temp2 <- unzip(temp)
  
  raleigh <- read_csv(temp2)
  
  save(raleigh, file = dat_path)
} else {
  load(dat_path)
}
```

# explore

I guess that the `time` variable is measured in seconds of the day

```{r}
raleigh$time %>% range(na.rm = TRUE)
# Time differences in secs
# [1]     1 86399
```

# Use loess.

```{r}
day_sec <- 60*60*24


# bin stops per minute
by_minute <- 
  raleigh %>%  
  mutate(mins = as_double(time) %/% 60,
         year = year(date)) %>% 
  group_by(mins, year) %>% 
  count() %>% 
  # na.omit() %>%
  # strange measurements at 0
  # probably tecnical issue
  filter(mins > 0) %>% 
  group_by(mins) %>% 
  summarise(n = mean(n))

# check
span <- 1/30
by_minute %>% 
  ggplot(aes(x = mins, y = n)) +
  stat_smooth(method = "loess", span = span) +
  geom_point(shape = ".") 

# smooth 
smooth_obj <- 
  by_minute %>% 
  {loess(formula = "n ~ mins",
         data = .,
         span = span)}


# prediction
preds <- 
  predict(smooth_obj, newdata = by_minute, se = T)
  
min_smooth <- 
  preds %>% 
  # fitted values and interval
  {tibble(mins = names(.$fit) %>% as.numeric(),
          fitted = .$fit,
          se = .$se)} %>% 
  # add back observed values
  full_join(by_minute, by = "mins") %>% 
  # interval at 2s
  mutate(lower_bound = fitted - (10*se),
         upper_bound = fitted + (10*se))

# check
min_smooth %>% 
  ggplot(aes(x = mins, y = n)) +
  geom_line(aes(y = fitted)) +
  geom_point(shape = ".")


# try ribbon
med_y <- mean(min_smooth$n)

min_smooth %>% 
  ggplot(aes(x = mins)) +
  geom_hline(yintercept = med_y,
             colour = "#B63A82") +
  geom_ribbon(aes(ymin = lower_bound,
                  ymax = upper_bound),
             fill = "grey80",
             colour = "#27A6D3") +
  geom_point(aes(y = n),
             shape = ".") +
  ylim(0, NA) +
  theme_minimal()

# ribbon 
fill_up <- "#3752C3"
min_smooth %>% 
  ggplot(aes(x = mins)) +
  geom_ribbon(aes(ymax = fitted,
                  ymin = med_y),
              fill = fill_up) +
  geom_point(aes(y = n),
             colour = "grey70",
             shape = ".") +
  theme_minimal()
```



# use coordinates
Only way that I know to plot an arc.

```{r}
# to coordinates
# y_from <- c(0, min_smooth$n %>% max)
# y_to <- c(70, 20)
# x_to <- c(30, -30)
# x_from <- range(min_smooth$mins)
# 
# to_plot <- 
#   min_smooth %>% 
#   mutate(mins = rescale(mins, to = x_to, from = x_from),
#          n = rescale(n, to = y_to, from = y_from),
#          fitted = rescale(fitted, to = y_to, from = y_from),
#          lower_bound = rescale(lower_bound, to = y_to, from = y_from),
#          upper_bound = rescale(upper_bound, to = y_to, from = y_from))
#   

to_plot <- min_smooth
```

# set parameters

```{r}
# save params again
med_y <- to_plot$n %>% mean
fill_down <- "#E97E00" # "#B63A82"
fill_up <- "#263A89" # "#3752C3"
ytop <- to_plot$n %>% max()
ylow <- to_plot$n %>% min()
ridge_width <- to_plot$fitted %>% range() %>% {(.[2] - .[1])/12}
grid_at <- c(20, 40, 60) %>% rescale(to = y_to, from = y_from)
annos_color <- "grey70"
text_color <- "#98F0D8"
mins_range <- range(to_plot$mins)
```

# test snake plot

```{r}
to_plot %>% 
  ggplot(aes(x = mins)) +
  geom_hline(yintercept = med_y,
             colour = "#B63A82") +
  geom_ribbon(aes(ymin = lower_bound,
                  ymax = upper_bound),
             fill = "grey80",
             colour = "#27A6D3") +
  geom_point(aes(y = n),
             shape = ".") +
  # ylim(0, NA) +
  theme_minimal() 
  # coord_map(projection = "azequidistant", orientation = c(90, -45, 225))

```

# Try trans_new()

```{r}
# 
# tst <- max(min_smooth$mins)
# 
# radial <- 
#   ggforce::radial_trans(r.range = c(-60, min_smooth$n %>% max()),
#                         a.range = range(tst*1/2, tst*8))
# 
# min_rad <- 
#   min_smooth %>% 
#   {radial$transform(#r = rep(mean(.$n), length(.$n)),
#     r = .$n,
#                       a = .$mins)}
# 
# ggplot(min_smooth, aes(x = mins, y = fitted)) +
#   geom_point()
# 
# ggplot(min_rad, aes(x = x, y = y)) +
#   geom_point() 

```

# Draft plot

```{r}
p <- 
  to_plot %>% 
  ggplot(aes(x = mins)) +
  lims(y = range(to_plot$n) + c(-2, +2),
       x = range(to_plot$mins) + c(-100, +100)) +
  theme_void()

p
```

# Add bars

```{r}
# add bars
add_bar <- function(p, at) {
  p <- 
    p + 
    annotate(geom = "rect",
             xmin = mins_range[1],
             xmax = mins_range[2],
             ymin = case_when(at >= 0 ~ Inf,
                              TRUE ~ -Inf),
             ymax = med_y + at,
             fill = case_when(at >= 0 ~ fill_up,
                              TRUE ~ fill_down),
             alpha = .3)
  return(p)  
} 

ridge_at <- c(-0.001, (-6:5)*ridge_width)

p1 <- p
for(i in ridge_at)  p1 <- add_bar(p1, i)

p1
```

# add snake


```{r}
p2 <-
  p1 +
  geom_ribbon(aes(ymin = -Inf,
                  ymax = lower_bound),
              fill = "white") +
  geom_ribbon(aes(ymin = upper_bound,
                  ymax = Inf),
              fill = "white") #+
  # geom_line(aes(y = upper_bound), colour = text_color) +
  # geom_line(aes(y = lower_bound), colour = text_color)
  
p2
```

# Add x guide / arrow

```{r}
p3 <- 
  p2 +
  geom_ribbon(aes(ymin = -Inf,
                  ymax = case_when(lower_bound >= med_y ~ med_y,
                                   TRUE ~ lower_bound)),
              fill = "white",
              colour = "#DE1288") +
  geom_ribbon(aes(ymin = +Inf,
                  ymax = case_when(upper_bound < med_y ~ med_y,
                                   TRUE ~ upper_bound)),
              fill = "white",
              colour = "#DE1288") +
  geom_line(aes(y = upper_bound),
            colour = "grey50") +
  geom_line(aes(y = lower_bound),
            colour = "grey50",
            size = 1.2)
              
  
p3  
```

# Save SVG

```{r}
svglite::svglite("plots/2-12-open-policing-snake.svg")
p3 + theme(aspect.ratio = .4)
dev.off()
```


```{r}
p <- 
to_plot %>% 
  ggplot(aes(x = mins,
             ymax = fitted)) +
  geom_ribbon(data = . %>%
                filter(fitted > med_y),
              ymin = med_y,
              fill = fill_up,
              # colour = colour,
              alpha = .2) +
  geom_ribbon(data = . %>%
                filter(fitted <= med_y),
              ymin = med_y,
              fill = fill_down,
              # colour = colour,
              alpha = .2)  +
  coord_map(projection = "azequidistant", orientation = c(90, -45, 225)) +
  # theme_minimal()
  theme_void()


p
```

```{r}

```


# add gradient

```{r}
up_gradient <- function(roll, p = p) {
  p <- p + 
    geom_ribbon(data = . %>%
                  filter(fitted > med_y + ridge_width*roll),
                aes(ymin = med_y + ridge_width*roll),
                fill = fill_up,
                alpha = .3)
  return(p)
}


down_gradient <- function(roll, p = p) {
  p <- p + 
    geom_ribbon(data = . %>%
                  filter(fitted < med_y - ridge_width*roll),
                aes(ymin = med_y - ridge_width*roll),
                fill = fill_down,
                alpha = .3)
  return(p)
}

p2 <- up_gradient(1, p)
p2 <- up_gradient(2, p2)
p2 <- up_gradient(3, p2)
p2 <- up_gradient(4, p2)
p2 <- up_gradient(5, p2)
p2 <- up_gradient(6, p2)
p2 <- down_gradient(1, p2)
p2 <- down_gradient(2, p2)
p2 <- down_gradient(3, p2)
p2 <- down_gradient(4, p2)
p2 <- down_gradient(5, p2)

p2
```

# cover ribbon residues

```{r}
# Cover up ribbon residues
p3 <- 
  p2 +
  geom_ribbon(data = . %>% 
                mutate(upper_bound = case_when(fitted < med_y ~ fitted,
                                               TRUE ~ med_y)),
              aes(ymax = upper_bound,
                  ymin = ylow),
              fill = "white",
              colour = NA) +
  geom_ribbon(data = . %>% 
                mutate(lower_bound = case_when(fitted > med_y ~ fitted,
                                               TRUE ~ med_y)),
              aes(ymax = ytop,
                  ymin = lower_bound),
              fill = "white",
              colour = NA) 
  
p3

```

# add grid

```{r}

add_lines <- function(at, p) {
  p <- 
    p +
    geom_hline(yintercept = at,
               lty = 2,
               size = .1,
               colour = annos_color)

  return(p)
}

p4 <- p3
for(i in grid_at) p4 <- add_lines(at = i, p = p4)
p4


```


# add points

```{r}
p5 <- 
  p4 +
  geom_point(aes(y = n),
             colour = "grey70",
             # shape = ".",
             size = .2,
             alpha = .5) 
```

# add y guide

```{r}
# add y guide
p6 <- 
  p5 + 
  geom_text(data = tibble(x = max(to_plot$mins) + 3,
                          y = grid_at),
            aes(x = x,
                y = y,
                label = y %>%
                  rescale(to = y_from, from = y_to) %>% 
                  round(1)),
            size = 5,
            nudge_x = 1,
            hjust = 1,
            vjust = .5,
            angle = max(to_plot$mins),
            colour = annos_color,
            inherit.aes = F) 

p6
```

# add arrow

```{r}
p7 <- 
  p6 + 
  geom_line(data = tibble(x = c(x_to[1] + 1.5, x_to[2] - 7),
                          y = med_y),
            aes(x = x, y = y),
            size = 1.2,
            arrow = arrow(ends = "first",
                          length = unit(3.4, "mm"),
                          type = "closed"),
            colour = text_color,
            inherit.aes = FALSE)

p7
```

# add minutes guide

```{r}
hrs <- seq(0, 24, by = 4) * 60
hrs_guides <- 
  hrs %>% 
  {tibble(x = rescale(., to = x_to, from = x_from),
          label = (./60))}


p8 <- 
  p7 +
  geom_text(data = hrs_guides,
            mapping = aes(x = x,
                          y = med_y,
                          label = paste0(label, "h"),
                          angle = x),
            size = 3,
            hjust = 0,
            vjust = 1,
            nudge_y = .5,
            colour = "#44D4DC", #"#27A6D3",
            fontface = "bold",
            inherit.aes = FALSE)

p8
```

# title and annos

```{r}
p9 <- 
  p8 +
  labs(title = "Average Police Stops Per Minute of the Day",
       subtitle = str_wrap("Recorded in Raleigh, NC., between
                           2002 and 2015. These data are gathered and
                           maintained by the
                           Stanford Open Policing Project.", 30),
       caption = paste0("Data from Stanford Open Policing Project.\n",
                        "Plot inspired to the design invented by ",
                        "Nadieh Bremer and Zan Armstrong for the ",
                        "Baby Spike article in Scientific American.\n",
                        "Plot done by @othomn in ggplot2.")) +
  theme(plot.title = element_text(family = "sans",
                             size = 12,
                             colour = "grey30", 
                             hjust = 1),
        plot.subtitle = element_text(family = "sans",
                             size = 10,
                             colour = "grey50", 
                             hjust = 1),
        plot.caption = element_text(family = "sans",
                                    face = "italic",
                                    size = 10,
                                    colour = "grey50", 
                                    hjust = 0))

p9

p10 <- 
  p9 +
  annotate(geom = "text",
           x = (5.5*60) %>% rescale(to = x_to, from = x_from),
           y = max(to_plot$fitted) + 2,
           hjust = 0,
           vjust = 1,
           label = str_wrap("I can imagine why fewer cars get
                            stopped by the police
                            at these hours of the day...", 20),
           family = "sans",
           size = 3,
           colour = "grey50", 
           lineheight = .9) + 
  annotate(geom = "text",
           x = (18.5*60) %>% rescale(to = x_to, from = x_from),
           y = max(to_plot$fitted) - 5,
           hjust = 0,
           vjust = 1,
           label = str_wrap("...but not why also at these hours.
                            Dinner time?", 20),
           family = "sans",
           size = 3,
           colour = "grey50", 
           lineheight = .9) 
  
p10

```

# Save SVG

```{r}
svglite::svglite("plots/2-12-open-policing-arc.svg")
p10
dev.off()
```

