---
title: "Open Policing - Raleigh"
author: "Otho"
date:  "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(lubridate)
library(tibbletime)
library(grid)

# weekdays in english
Sys.setlocale("LC_TIME", "en_US.UTF8")
```

# get data

```{r}
bike_url <- paste0("https://raw.githubusercontent.com/rfordatascience/",
                   "tidytuesday/master/data/2019/2019-04-02/bike_traffic.csv")
bike_path <- "data/2-14-seattle-bike.Rdata"

if(!file.exists(bike_path)) {
  bike_traffic <- readr::read_csv()
   
  save(bike_traffic, file = "data/2-14-seattle-bike.Rdata")
} else {
  load("data/2-14-seattle-bike.Rdata")
}
```  

# wrangle data

Fix time

```{r}
bike_traffic <- 
  bike_traffic %>% 
  mutate(date = mdy_hms(date))
```

Explore

```{r, eval = FALSE}
# explore -----------------------------------------------------------------
bike_traffic %>% 
  pull(crossing) %>% 
  unique()

bike_traffic %>% 
  mutate(year_day = yday(.data$date),
         year = year(.data$date)) %>% 
  group_by(year, year_day, crossing) %>% 
  summarise(bike_count = sum(bike_count, na.rm = TRUE))

bike_traffic %>% 
  ggplot(aes(x = bike_count)) +
  geom_histogram() +
  scale_x_log10() +
  facet_grid(crossing ~ .)


bike_traffic %>% 
  mutate(week_day = wday(.data$date),
         week = week(.data$date), 
         year = year(.data$date)) %>% 
  # group_by(year, year_day, crossing) %>% 
  filter(year > 2013,
         year < 2019) #arrange(desc(bike_count)) %>%  top_n(n = 10)
```

After exploring the data, I decided to go for a poster with 4 heavily faceted panel, for two reasons:

- I thought that it was interesting to show bike counts for each crossing across every day of a year. So one could see the differences between working days and weekends and between the warm and cold season.
- I wanted to practice poster-like layouts and heavily faceted plots, 

I decided to use 2017 because measurements seem most complete during this year.

First, I had to decide what to show on a single plots

Use rolling mean.


```{r}
# smooth mean on a window of three measurements
roll_mean <- rollify(window = 3, .f = mean)

to_plot <- 
  bike_traffic %>% 
  # no need, but I wanted to use something other than "date"
  # which is generic and it is a function
  rename(time_stamp = "date") %>% 
  # each crossing measures bikes in different direction separately
  # sum all the values to get counts by crossing an by hours
  group_by(crossing, time_stamp) %>% 
  summarise(bike_count = sum(bike_count)) %>% 
  ungroup() %>% 
  # Not necessary, but some of this column might be useful later
  mutate(year = year(time_stamp),
         year_day = yday(time_stamp),
         week_day = wday(time_stamp),
         day_hour = hour(time_stamp),
         date_day = round_date(time_stamp, unit = "day")) %>% 
  # seems like measurements are most complete in this year
  filter(year == 2017) %>%
  # filter(year_day > 45,year_day < 50) %>% 
  group_by(crossing) %>%
  arrange(year_day, day_hour) %>% 
  mutate(s_mean = roll_mean(bike_count)) %>% 
  ungroup()
```

# Store all plots in a list

I want to control the position of every plot individually
Thus, I plot every day in separately and put everything in a list.


```{r}
# First some parameters
# day 230 has unusually high counts and flattens everything else
ytop <-
  to_plot %>% 
  filter(year_day != 230) %>% 
  pull(s_mean) %>% max(na.rm = TRUE)

plot_bikes <- function(day_in)
{
  p <- to_plot %>%
    filter(yday(time_stamp) == yday(day_in)) %>% 
    ggplot(aes(x = day_hour,
               fill = crossing,
               colour = crossing)) +
    geom_ribbon(aes(ymin = 0,
                    ymax = s_mean),
                alpha = .2) +
    guides(colour = FALSE,
           fill = FALSE) +
    lims(y = c(0, ytop)) +
    theme_void() +
    theme(plot.margin = margin(0, 10, 0, 10))
  
  return(p)
}

# one plot per day
p_list <- 
  to_plot$time_stamp %>%
  round_date(unit = "day") %>%
  unique() %>% 
  {subset(., year(.) == 2017)} %>% 
  map(plot_bikes)


save(p_list, file = "data/2-14-seattle-bike-PLOTS.Rdata")
```

# Set the grid

To place all plots in the grid programmatically, first you have to set some parameter.

```{r}
length(p_list)
# 365

# how many rows per block (4 blocks in total)
# +1 to fit the remainder
length(p_list)%/%(7*4) + 1
# 14

# how many plots per block (approx)
length(p_list)%/%4
# 91

# rows per block
block_rows <- 14
block_cols <- 7
# 112 cells per block

14*7*4
# 392 cells for 365 plots
# 27 cells too much - no big deal:
# a bit of empty space looks ok at the end
 

# outer margin left and right
m_side <- .14
# outer margin up and down
m_tb <- .16
# inner margin left and right
m_small_side <- .04
# inner margin top and bottom
m_small_tb <- .05

# must fit a group of 8 plots in half table width
# the first space is for the date annotation
# the other 7 are for the plots
p_width <- (.5 - m_side - m_small_side)/8 

# must 13 plots in half table height
p_height <- (.5 - m_tb - m_small_tb)/13


# background color
bg_col <- "#F1F1F0" #"#F2F2EF"#"#F6F6EC"
```

Place the coordinates of each plot in a dataframe,
then you will loop on that dataframe with pmap()

```{r}

# x for third and fourth block are a repeat of this:
x_firstblock <-
  seq(from = m_side,
      to = .5 - m_small_side,
      length.out = 9)[2:(block_cols + 1)] %>% 
  rep(block_rows) 

x_secondblock <-
  c(seq(from = .5 + m_small_side,
        to = 1 - m_side,
        length.out = 9)[2:(block_cols + 1)] %>% 
      rep(block_rows))

# y for second and fourth block are a repeat of first and third
y_firstblock <- 
  seq(from = 1 - m_tb,
      to = .5 + m_small_tb,
      length.out = block_rows) %>%
  rep(each = block_cols) %>% 
  {c(., .)}

y_thirdblock <- 
  c(seq(from = .5 - m_small_tb, to = m_tb,
        length.out = block_rows) %>% 
      rep(each = block_cols) %>% 
      {c(., .)})



# in a data frame?
p_tibble <- 
  tibble(p = p_list,
         x = c(x_firstblock, x_secondblock) %>% 
           # # third and fourth blocks (bottom)
           {c(.,.)} %>% 
           # remove extra positions
           .[1:length(p_list)],
         y = c(y_firstblock, y_thirdblock) %>% 
           .[1:length(p_list)],
         width = p_width,
         height = p_height)


# define a function to plot each plot on its position
# in the grid
plot_to_vp <- function(p, x, y, width, height) {
  print(p,
        vp = viewport(x = x, y = y, width = width, height = height))
  return(NULL)
}


# Very slow output on the screen, don't know why  ------------------------

# grid.newpage()
# 
# p_tibble %>% pmap(plot_to_vp)


# much faster on SVG ------------------------------------------------------

# function to add weeknames
# w_names <- 
#   tibble(x = seq(from = m_side, to = .5 - m_small_side, length.out = 9)[1:block_cols] %>% 
#            c(seq(from = .5 + m_small_side, to = 1 - m_side, length.out = 9)[1:block_cols]),
#          label = c("", "Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat") %>% 
#            rep(2))
# 
# add_weeknames <- function(x, label) {
#   grid.text(label = label,
#             x = x,
#             y = 1 - m_tb*.95,
#             vjust = 0,
#             gp = gpar(col = "grey20",
#                       fontsize = 18,
#                       fontface = "italic",
#                       fontfamily = "Times New Roman"))
#   return(NULL)
# }
#   
# w_names %>% pmap(add_weeknames)


svglite::svglite(file = "plots/2-14-seattle-bikes-draft-grid-2.svg",
                 height = 33.1,
                 width = 23.4 )
grid.newpage()
# background
grid.rect(gp = gpar(fill = bg_col))
# Title
grid.text("365 Days Cycling in Seattle",
          x = m_side,
          y = 1 - m_tb/2,
          gp = gpar(col = "grey20",
                    fontsize = 60,
                    fontface = "italic",
                    fontfamily = "Times New Roman"),
          hjust = 0,
          vjust = 0)
# subtitle
grid.text(str_wrap("Bicycle traffic crossing 7 detection points in Seattle in 2017.") %>% 
            paste(str_wrap(" Also an exercise on making heavily facetted plots with ggplot2 and
                   grid by Otho Mantegazza.",
                   width = 100), sep = "\n"),
          x = m_side + .02, 
          y = 1 - m_tb*.72,
          gp = gpar(col = "grey20",
                    fontsize = 27,
                    fontface = "italic",
                    fontfamily = "Times New Roman",
                    lineheight = .87),
          hjust = 0,
          vjust = 0)
# mid line
grid.lines(x = unit(c(m_side*1.5, 1 - m_side*1.5), "npc"),
           y = unit(c(.5, .5), "npc"))
# week names
w_names %>% pmap(add_weeknames)
# plots
p_tibble %>% pmap(plot_to_vp)
# caption
grid.text(label = "Data from the Seattle Department of Transportation\nplot by @othomn",
          x = 1 - m_side, 
          y = m_tb*.5,
          gp = gpar(col = "grey20",
                    fontsize = 27,
                    fontface = "italic",
                    fontfamily = "Times New Roman",
                    lineheight = .87),
          hjust = 1,
          vjust = 0)
dev.off()

png(filename = "plots/2-14-seattle-bikes-draft-grid-2.png", 
    height = 33.1,
    width = 23.4,
    units = "in",
    res = 300)
grid.newpage()
# background
grid.rect(gp = gpar(fill = bg_col))
# Title
grid.text("365 Days Cycling in Seattle",
          x = m_side,
          y = 1 - m_tb/2,
          gp = gpar(col = "grey20",
                    fontsize = 60,
                    fontface = "italic",
                    fontfamily = "Times New Roman"),
          hjust = 0,
          vjust = 0)
# subtitle
grid.text(str_wrap("Bicycle traffic crossing 7 detection points in Seattle in 2017.") %>% 
            paste(str_wrap(" Also an exercise on making heavily facetted plots with ggplot2 and
                   grid by Otho Mantegazza.",
                           width = 100), sep = "\n"),
          x = m_side + .02, 
          y = 1 - m_tb*.72,
          gp = gpar(col = "grey20",
                    fontsize = 27,
                    fontface = "italic",
                    fontfamily = "Times New Roman",
                    lineheight = .87),
          hjust = 0,
          vjust = 0)
# mid line
grid.lines(x = unit(c(m_side*1.5, 1 - m_side*1.5), "npc"),
           y = unit(c(.5, .5), "npc"))
# week names
w_names %>% pmap(add_weeknames)
# plots
p_tibble %>% pmap(plot_to_vp)
# caption
grid.text(label = "Data from the Seattle Department of Transportation\nplot by @othomn",
          x = 1 - m_side, 
          y = m_tb*.5,
          gp = gpar(col = "grey20",
                    fontsize = 27,
                    fontface = "italic",
                    fontfamily = "Times New Roman",
                    lineheight = .87),
          hjust = 1,
          vjust = 0)
dev.off()

```

